Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-05-25 analysis/paper/paper.qmd
revise the method section

Only faunal assemblages identified to at least the family level were considered in this study to ensure consistent comparison across time sequences. Deer antlers detached from the cranium were excluded because they were less likely to be related to dietary practices. To obtain a generalized pattern of animal consumption, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their consumption [@Mckechnie2016; @Vanderwarker2010]. In this study, ubiquity values are expressed as the percentage of units containing a given taxon relative to the total number of units per cultural layer. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012]. MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages. Despite criticism for potentially overestimating the number of animals, NISP better captures variation in species abundance with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison across animal classes, the normalized NISP were used to avoid biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin could be a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This analysis can provide insights into how Indigenous people utilized deer in response to possible high economic demand when driven by their involvement in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty parts (e.g., upper limbs), mid-meaty parts (e.g., ribs), and less meaty parts (e.g., skulls and lower limbs) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded to avoid overrepresentation, given their much higher preservation rate and quantities compared to other parts. Despite having less meat, lower limbs and skulls are often valued in many cultures for their marrow and soft tissues, such as the tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as an additional source of nutrition in later interpretations. Similarly, the degree of fragmentation could indicate marrow or grease extraction [@Lupo1997], and was calculated by dividing the Minimum Number of Elements (MNE) by NISP. In this approach, lower values represent a higher degree of fragmentation, and higher values indicate a lower degree of fragmentation.

>>>>

Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012]. MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages. Despite criticism for potentially overestimating the number of animals, NISP can better capture variation in species abundance with less effect from sample size [@Grayson1984; @Lyman2008]. Bone fragments identified to the class level (e.g., Mammalia, Aves) were included in the overall NISP count to reflect the full scope of vertebrate remains. However, to enable more detailed taxonomic comparisons, indeterminate Mammalia specimens were excluded from further analysis to focus on bones identified to at least the family level. Deer antlers detached from the cranium were not considered in the quantification because they were less likely to be associated with dietary practices. To reduce biases resulting from differential number and recovery of skeletal elements among animal classes, NISP values were normalized by dividing the NISP by the average number of skeletal elements in a complete individual of that class [@Grayson2004].

To identify generalized patterns of animal consumption, measures of taxonomic abundance and ubiquity were applied. Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their use [@Mckechnie2016; @Vanderwarker2010]. In this study, ubiquity values are expressed as the percentage of units in which a given taxon is present relative to the total number of units per cultural layer. Given that mammals were the most commonly found animal resource, mammal bones identified to the family level were categorized individually for further discussion and compared in terms of their ubiquity with other animal classes.

Since deerskin may have been a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This analysis can provide insights into how Indigenous people utilized deer in response to possible high economic demand when driven by their involvement in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty parts (e.g., upper limbs), mid-meaty parts (e.g., ribs), and less meaty parts (e.g., skulls and lower limbs) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded to avoid overrepresentation, given their much higher preservation rate and quantities compared to other parts. Despite having less meat, lower limbs and skulls are often valued in many cultures for their marrow and soft tissues, such as the tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as an additional source of nutrition in later interpretations. Similarly, the degree of fragmentation could indicate marrow or grease extraction [@Lupo1997], and was calculated by dividing the Minimum Number of Elements (MNE) by NISP. In this approach, lower values represent a higher degree of fragmentation, and higher values indicate a lower degree of fragmentation.
